Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2017-11-24 docs/manuscript.tex
MS updated with some results

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief \citep{DataThief}, GraphClick \citep{GraphClick}, WebPlotDigitizer \citep{WebPlotDigitizer}), these tools are not designed specifically for meta-analysis. 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types
Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required, that differs for different plots types. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata
Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations more laborious, as the information has to be added later, in most cases using different software.
%% 3 - in same statistical software and high throughput
Finally, existing programs do not import a set of images and allow the user to systematically work through them. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. In essence, existing software does not provide an optimized research pipeline to facilitate data extraction, editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013, Parker2016}.

@@

@@

  Across plot types we found no evidence for any inter-observer variability in digitisations for the mean (Unadjusted ICC = 0, 95 CI = 0 to 0.029, p = 1; Adjusted ICC = 0, 95 CI = 0 to 0.033, p = 1), standard deviation (Unadjusted ICC = ; Adjusted ICC = ) or correlation coefficient (Unadjusted ICC = ; Adjusted ICC = )

>>>>

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief \citep{DataThief}, GraphClick \citep{GraphClick}, WebPlotDigitizer \citep{WebPlotDigitizer}), these tools are not designed specifically for meta-analysis. 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types

Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required, that differs for different plots types. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata

Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations more laborious, as the information has to be added later, in most cases using different software.
%% 3 - in same statistical software and high throughput

Finally, existing programs do not import a set of images and allow the user to systematically work through them. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. In essence, existing software does not provide an optimized research pipeline to facilitate data extraction, editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013, Parker2016}.

@@

\subsection{Testing the accuracy of digitisations}
To test how accurate \pkg{metaDigitise} is at matching points to their true values, we generated four random scatterplots and tested how well the digitised point in each scatterplot matches the actual, true point generated from simulations. We evaluated the accuracy using a Pearson's correlation coefficient between the digitised data from \pkg{metaDigitise} and the true known data value from the simulated data.

@@

Across plot types we found no evidence for any inter-observer variability in digitisations for the mean (Unadjusted ICC = 0, 95\% CI = 0 to 0.029, p = 1; Adjusted ICC = 0, 95\% CI = 0 to 0.033, p = 1), standard deviation (Unadjusted ICC = 0, 95\% CI = 0 to 0.033, p = 0.5; Adjusted ICC = 0.002, 95\% CI = 0 to 0.032, p = 1) or correlation coefficient (Unadjusted ICC = 0.053, 95\% CI = 0 to 0.296, p = 0.377; Adjusted ICC = 0.048, 95\% CI = 0 to 0.3, p = 0.374). Overall, there were very small differences between digitised values and the true values only deviating, on average 3.34\% (mean = 0.02\%, sd = 5\%, r = 5\%) for all three summary statistics.
%% Note that unadjusted and adjusted values come from rptR and the deviation of percent are the intercept values (*100) for mean, sd and r. I've averaged these three for 3.34%

Digitised data was essentially perfectly correlated with the true simulated data for both the x-variable (r = 0.9999915, t = 2137.4, df = 78, p-value < 2.2e-16) and y-variable (r = 0.9999892, t = 1897.8, df = 78, p-value < 2.2e-16) in scatterplots.
